---
title: "Water Level for Nigeria"
subtitle: "Take-home Exercise 1"
author: "Yang Jingyuan"
execute: 
  warning: false
editor: visual
---

November 27, 2022

## Overview

This is take-home exercise 1.

## Setting the Scene

Water is an important resource to mankind. Clean and accessible water is critical to human health. It provides a healthy environment, a sustainable economy, reduces poverty and ensures peace and security. Yet over 40% of the global population does not have access to sufficient clean water. By 2025, 1.8 billion people will be living in countries or regions with absolute water scarcity, according to UN-Water. The lack of water poses a major threat to several sectors, including food security. Agriculture uses about 70% of the world's accessible freshwater.

Developing countries are most affected by water shortages and poor water quality. Up to 80% of illnesses in the developing world are linked to inadequate water and sanitation. Despite technological advancement, providing clean water to the rural community is still a major development issues in many countries globally, especially countries in the Africa continent.

To address the issue of providing clean and sustainable water supply to the rural community, a global [Water Point Data Exchange (WPdx)](https://www.waterpointdata.org/about/) project has been initiated. The main aim of this initiative is to collect water point related data from rural areas at the water point or small water scheme level and share the data via WPdx Data Repository, a cloud-based data library. What is so special of this project is that data are collected based on [WPDx Data Standard](https://www.waterpointdata.org/wp-content/uploads/2021/04/WPDx_Data_Standard.pdf).

## Objectives

Geospatial analytics hold tremendous potential to address complex problems facing society. In this study, you are tasked to apply appropriate global and local measures of spatial Association techniques to reveals the spatial patterns of Not Functional water points. For the purpose of this study, Nigeria will be used as the study country.

## The Data

### Apstial data

For the purpose of this assignment, data from [WPdx Global Data Repositories](https://www.waterpointdata.org/access-data/) will be used. There are two versions of the data. They are: WPdx-Basic and WPdx+. You are required to use WPdx+ data set.

### Geospatial data

Nigeria Level-2 Administrative Boundary (also known as Local Government Area) polygon features GIS data will be used in this take-home exercise. The data can be downloaded either from The [Humanitarian Data Exchange](https://data.humdata.org/) portal or [geoBoundaries](https://www.geoboundaries.org/).

## The Task

The specific tasks of this take-home exercise are as follows:

-   Using appropriate sf method, import the shapefile into R and save it in a simple feature data frame format. Note that there are three Projected Coordinate Systems of Nigeria, they are: EPSG: 26391, 26392, and 26303. You can use any one of them.

-   Using appropriate tidyr and dplyr methods, derive the proportion of functional and non-functional water point at LGA level.

-   Combining the geospatial and aspatial data frame into simple feature data frame.

-   Performing outliers/clusters analysis by using appropriate local measures of spatial association methods.

-   Performing hotspot areas analysis by using appropriate local measures of spatial association methods.

### Thematic Mapping

-   Plot maps to show the spatial distribution of functional and non-functional water point rate at LGA level by using appropriate thematic mapping technique provided by tmap package.

### Analytical Mapping

-   Plot hotspot areas and outliers/clusters maps of functional and non0functional water point rate at LGA level by using appropriate thematic mapping technique provided by tmap package.

## Getting Started

```{r}
pacman::p_load(sf, tidyverse, tmap, spdep, funModeling)
```

## Importing Data into R

In this in-class data, two geospatial data sets will be used, they are:

-   geo_export

-   nga_admbnda_adm2_osgof_20190417

### Importing Geospatial Data

First, we are going to import the water point geospatial data (i.e. geo_export) by using the code chunk below.

```{r}
#| eval: false
wp <- st_read(dsn = "data/geospatial", 
              layer = "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

Things to learn from the code chunk above:

-   `st_read()` of **sf** package is used to import *geo_export* shapefile into R environment and save the imported geospatial data into simple feature data table.

-   `filter()` of dplyr package is used to extract water point records of Nigeria.

> Be warned: Avoid performing transformation if you plan to use `st_intersects()` of **sf** package in the later stage of the geoprocessing. This is because `st_intersects()` only works correctly if the geospatial data are in geographic coordinate system (i.e. wgs84)

Next, `write_rds()` of readr package is used to save the extracted sf data table (i.e. wp) into an output file in rds data format. The output file is called *wp_nga.rds* and it is saved in *geodata* sub-folder.

```{r}
#| eval: false
write_rds(wp, "data/geospatial/wp_nga.rds")
```

### Importing Nigeria LGA boundary data

Now, we are going to import the LGA boundary data into R environment by using the code chunk below.

```{r}
#| eval: false
nga <- st_read(dsn = "data/geospatial",
               layer = "nga_admbnda_adm2_osgof_20190417",
               crs = 4326)
```

Thing to learn from the code chunk above.

-   `st_read()` of **sf** package is used to import *nga_admbnda_adm2_osgof_20190417* shapefile into R environment and save the imported geospatial data into simple feature data table.

## Data Wrangling

### Recoding NA values into string

In the code chunk below, `replace_na()` is used to recode all the *NA* values in *status_cle* field into *Unknown*.

```{r}
#| eval: false
wp_nga <- read_rds("data/geospatial/wp_nga.rds") %>%
  mutate(status_cle = replace_na(status_cle, "Unknown"))
```

### EDA

In the code chunk below, `freq()` of **funModeling** package is used to display the distribution of *status_cle* field in *wp_nga*.

```{r}
#| eval: false
freq(data=wp_nga, 
     input = 'status_cle')
```

## Extracting Water Point Data

In this section, we will extract the water point records by using classes in *status_cle* field.

### Extracting funtional water point

In the code chunk below, `filter()` of dplyr is used to select functional water points.

```{r}
#| eval: false
wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional", 
             "Functional but not in use",
             "Functional but needs repair"))
```

```{r}
#| eval: false
freq(data=wpt_functional, 
     input = 'status_cle')
```

### Extracting non-funtional water point

In the code chunk below, `filter()` of dplyr is used to select non-functional water points.

```{r}
#| eval: false
wpt_nonfunctional <- wp_nga %>%
  filter(status_cle %in%
           c("Abandoned/Decommissioned", 
             "Abandoned",
             "Non-Functional",
             "Non functional due to dry season",
             "Non-Functional due to dry season"))
```

```{r}
#| eval: false
freq(data=wpt_nonfunctional, 
     input = 'status_cle')
```

### Extracting water point with Unknown class

In the code chunk below, `filter()` of dplyr is used to select water points with unknown status.

```{r}
#| eval: false
wpt_unknown <- wp_nga %>%
  filter(status_cle == "Unknown")
```

## Performing Point-in-Polygon Count

```{r}
#| eval: false
nga_wp <- nga %>% 
  mutate(`total wpt` = lengths(
    st_intersects(nga, wp_nga))) %>%
  mutate(`wpt functional` = lengths(
    st_intersects(nga, wpt_functional))) %>%
  mutate(`wpt non-functional` = lengths(
    st_intersects(nga, wpt_nonfunctional))) %>%
  mutate(`wpt unknown` = lengths(
    st_intersects(nga, wpt_unknown)))
```

## Saving the Analytical Data Table

```{r}
#| eval: false
nga_wp <- nga_wp %>%
  mutate(pct_functional = `wpt functional`/`total wpt`) %>%
  mutate(`pct_non-functional` = `wpt non-functional`/`total wpt`) %>%
  select(3:4, 9:10, 18:23)
```

Things to learn from the code chunk above:

-   `mutate()` of **dplyr** package is used to derive two fields namely *pct_functional* and *pct_non-functional*.

-   to keep the file size small, `select()` of **dplyr** is used to retain only field 3,4,9,10, 18,19,20,21,22,and 23.

Now, you have the tidy sf data table subsequent analysis. We will save the sf data table into rds format.

```{r}
#| eval: false
write_rds(nga_wp, "data/geospatial/nga_wp.rds")
```

Before you end this section, please remember to delete away all the raw data. Notice that the only data file left is *nga_wp.rds* and it's file size is aroung 2.1MB.

## Visualising the spatial dsitribution of water points

```{r}
#| eval: false
nga_wp <- read_rds("data/geospatial/nga_wp.rds")
total <- qtm(nga_wp, "total wpt")
wp_functional <- qtm(nga_wp, "wpt functional")
wp_nonfunctional <- qtm(nga_wp, "wpt non-functional")
unknown <- qtm(nga_wp, "wpt unknown")

tmap_arrange(total, wp_functional, wp_nonfunctional, unknown, asp=1, ncol=2)
```

## Grading Criteria

This exercise will be graded by using the following criteria:

-   **Geospatial Data Wrangling (20 marks):** This is an important aspect of geospatial analytics. You will be assessed on your ability to employ appropriate R functions from various R packages specifically designed for modern data science such as readr, tidyverse (tidyr, dplyr, ggplot2), sf just to mention a few of them, to perform the entire geospatial data wrangling processes, including. This is not limited to data import, data extraction, data cleaning and data transformation. Besides assessing your ability to use the R functions, this criterion also includes your ability to clean and derive appropriate variables to meet the analysis need. (Warning: All data are like vast grassland full of land mines. Your job is to clear those mines and not to step on them).

-   **Geospatial Analysis (30 marks):** In this exercise, you are expected to use the appropriate thematic and analytics mapping techniques and R functions introduced in class to analysis the geospatial data prepared. You will be assessed on your ability to derive analytical maps by using appropriate rate mapping techniques.

-   **Geovisualisation (20 marks):** In this section, you will be assessed on your ability to communicate the complex spatial statistics results in business friendly visual representations. This course is geospatial centric, hence, it is important for you to demonstrate your competency in using appropriate geovisualisation techniques to reveal and communicate the findings of your analysis.

-   **Reproducibility (20 marks):** This is an important learning outcome of this exercise. You will be assessed on your ability to provide a comprehensive documentation of the analysis procedures in the form of code chunks of RMarkdown. It is important to note that it is not enough by merely providing the code chunk without any explanation on the purpose and R function(s) used.

-   **Bonus (10 marks):** Demonstrate your ability to employ methods beyond what you had learned in class to gain insights from the data. The methods used must be geospatial in nature.

## Submission Instructions

-   The write-up of the take-home exercise must be in [**distill**](https://rstudio.github.io/distill/) or [**blogdown**](https://github.com/rstudio/blogdown) format. You are required to publish the write-up on [**Netlify**](https://www.netlify.com/).

-   The R project of the take-home exercise must be pushed onto your [Github](https://github.com/) repository.

-   You are required to provide the links to Netlify service of the take-home exercise write-up and github repository on eLearn.

## Due Date

29th November 2022 (Tuesday), 11.59pm (midnight).

## Reference
