---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
November 18, 2022
---

## Overview

In this hands-on exercise, I learn how to import and wrangling geospatial data using appropriate R packages.

## Learning Objective

In this hands-on exercise, you will learn how to handle geospatial data in R by using **sf** package.

By the end of this hands-on exercise, you should acquire the following competencies:

-   installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/) packages into R environment,

-   importing geospatial data by using appropriate functions of **sf** package,

-   importing aspatial data by using appropriate function of **readr** package,

-   exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,

-   assigning or transforming coordinate systems by using using appropriate *sf* functions,

-   converting an aspatial data into a sf data frame by using appropriate function of **sf** package,

-   performing geoprocessing tasks by using appropriate functions of **sf** package,

-   performing data wrangling tasks by using appropriate functions of **dplyr** package and

-   performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

## Getting Started

The code chunk below install and load sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```
